\chapter{Acknowledgments}
Throughout the three years that I have spent working on this thesis I had support from many people, who I would like to thank here.

First of, I would like to thank the entire Albert-Einstein-Institute for their support, the opportunities they have provided me with, and the friendly working environment. The discussions I could partake in were always very fruitful. I specifically want to thank Bruce Allen, Alexander Harvey Nitz, and Badri Krishnan for allowing me to join the institute and the compact binary merger group. I also want to thank the Atlas computing team at the Albert-Einstein-Institute for their support, without whom most of the work presented in this thesis would not have been possible.

I am especially grateful for the guidance I have received by Alexander Harvey Nitz, who in most aspects has been my mentor. He suggested many of the research topics covered in this thesis but was also always open to new ideas and allowed me to experiment. His insights and tips have been invaluable to the completion of this thesis.

Another person who has greatly influenced my scientific evolution is Frank Ohme. He was my first point of contact with the Albert-Einstein-Institute and has always tried to enable my development wherever possible. His scientific input has greatly improved the quality of the works presented in this thesis.

I would also like to thank all co-authors of the various publications that have culminated in this thesis. These people are Bernd Br{\"u}gmann, Miriam Cabero, Zhoujian Cao, Collin D. Capano, Elena Cuoco, Tito Dal Canton, Rahul Dhurkunde, Zong-Kuan Guo, Eliu A. Huerta, Panagiotis Iosif, Shilpa Kastha, Sergey Klimenko, Alexandra E. Koloniari, Sumit Kumar, Chris Messenger, Tanmaya Mishra, Alexander H. Nitz, Paraskevi Nousi, Frank Ohme, Nikolaos Passalis, Zhixiang Ren, Francesco Salemi, Nikolaos Stergioulas, Anastasios Tefas, Gabriele Vedovato, He Wang, Yi-Fan Wang, Shichao Wu, and Ond{\v{r}}ej Zelenka.

A major contribution to the completion of this thesis have been my family and my close friends. Their moral and emotional support has been essential for my well being during this time. Most importantly I want to express my gratitude towards my mother Marie Therese Schäfer and my partner Nadine Speer, who were a pillar especially during difficult times. I also want to thank my father Thomas Eberhard Kurt Schäfer as well as Doris Maria Schulz, Helmger Hans Roderich Schotola, Felizitas Kaya, Horst Paul Hermann Schrödter, and Brigitta Beatrice Kuhr for support during all stages of my life.

I would further like to thank Bosch Hildesheim for allowing me to have a look into the non-academic work life and for interesting insights into computer vision applications of deep learning. I would like to especially thank Niklas Beuter, Hendrik Klose, and Christoph Dreißigacker, who made the internship possible. Furthermore, I would like to thank Lukas Enderich, who I closely worked with, as well as the entire closed loop learning project for the warm welcome and the open discussions.

Finally, I would like to thank the people who proofread the majority of this thesis. They significantly improved clarity, grammar, and spelling. Thank you Tobias Blenke, Christoph Dreißigacker, Lukas Enderich, Tom Junker, Alexander H. Nitz, Frank Ohme, Lina Schmitz, Nadine Speer, and Ond\v{r}ej Zelenka.
