{
\titleformat{\chapter}{\Huge\bfseries}{}{0pt}{}
\titlespacing{\chapter}{0pt}{-20pt}{0pt}

\chapter{Curriculum Vitae}
{\noindent\Large Marlin Benedikt Schäfer}
\vspace*{1cm}

\vspace*{-0.5cm}
\noindent
\begin{tabular}{ll}
\textbf{Date of birth} & 28th December 1995 \\
\textbf{Place of birth} & Hannover, Germany \\
\textbf{Email} & marlin.schaefer@aei.mpg.de
\end{tabular}

\vspace*{-0.5cm}
\section*{Education}
\vspace*{-0.5cm}
{\renewcommand{\arraystretch}{1.25}
\begin{tabularx}{\textwidth}{lX}
\textbf{2017 -- 2019} & MSc Physik -- Leibniz Universität Hannover\\
 & Master thesis\\
 &
 	{
 	\begin{tabularx}{\linewidth}{X}
 		Analysis of Gravitational-Wave Signals from Binary Neutron Star Mergers Using Machine Learning\\
 		Supervisor: Dr. Frank Ohme
 	\end{tabularx}
 	}\\
\textbf{2014 -- 2018} & BSc Physik -- Leibniz Universität Hannover\\
 & Bachelor thesis\\
 &
 	{\begin{tabularx}{\linewidth}{X}
 		 Massenbestimmung kompakter Binärsysteme durch Analyse von Gravitationswellen\\
 		 Determination of Component Masses from Compact Binary System by Analyzing Gravitational Waves \\
 		 Supervisor: Prof. Domenico Giulini
	\end{tabularx}}\\
\textbf{2006 -- 2014} & Abitur -- Kaiser-Wilhelm- und Ratsgymnasium Hannover
\end{tabularx}
}

\vspace*{-0.5cm}
\section*{Experience}
\vspace*{-0.5cm}
\begin{tabularx}{\textwidth}{lX}
\textbf{Nov. 2019 -- Dec. 2022} & Doctoral researcher -- Albert-Einstein-Institute Hannover\\
\textbf{Jun. 2022 -- Sep. 2022} & Research intern -- Robert Bosch GmbH Hildesheim\\
\textbf{Apr. 2019 -- Sep. 2019} & Assistant researcher -- Albert-Einstein-Institute Hannover\\
\textbf{Oct. 2016 -- Mar. 2019} & Tutor for theoretical physics -- Leibniz Universität Hannover\\
\textbf{Aug. \& Sep. 2016 -- 2018} & Tutor for mathematics -- Leibniz Universität Hannover
\end{tabularx}
}
