\chapter{List of Publications}\label{app:publications}
The research for this thesis has resulted in the following publications:
\begin{itemize}
	\item[\cite{Schafer:2020kor}] Marlin B. Schäfer, Frank Ohme, and Alexander H. Nitz, \href{https://doi.org/10.1103/PhysRevD.102.063015}{Detection of gravitational-wave signals from binary neutron star mergers using machine learning.} Phys. Rev. D 102.6 (2020)
	\item[\cite{Nitz:2020vym}] Alexander H. Nitz, Marlin B. Schäfer, and Tito Dal Canton, Forecasting \href{https://doi.org/10.3847/2041-8213/abbc10}{Gravitational-wave Merger Forecasting: Scenarios for the Early Detection and Localization of Compact-binary Mergers with Ground-based Observatories.} Astrophys. J. Lett. 902 (2020)
	\item[\cite{Nitz:2021uxj}] Alexander H. Nitz, Collin D. Capano, Sumit Kumar, Yi-Fan Wang, Shilpa Kastha, Marlin B. Schäfer, Rahul Dhurkunde and Miriam Cabero, \href{https://doi.org/10.3847/1538-4357/ac1c03}{3-OGC: Catalog of Gravitational Waves from Compact-binary Mergers.} Astrophys. J. 922.1 (2021)
	\item[\cite{Schafer:2021fea}] Marlin B. Schäfer and Ond\v{r}ej Zelenka, Alexander H. Nitz, Frank Ohme and Bernd Brügmann, \href{https://doi.org/10.1103/PhysRevD.105.043002}{Training strategies for deep learning gravitational-wave searches.} Phys. Rev. D 105.4 (2022)
	\item[\cite{Schafer:2021cml}] Marlin B. Schäfer and Alexander H. Nitz, \href{https://doi.org/10.1103/PhysRevD.105.043003}{From one to many: A deep learning coincident gravitational-wave search.} Phys. Rev. D 105.4 (2022)
	\item[\cite{Nitz:2021zwj}] Alexander H. Nitz, Sumit Kumar, Yi-Fan Wang, Shilpa Kastha, Shichao Wu, Marlin B. Schäfer, Rahul Dhurkunde and Collin D. Capano, \href{http://arxiv.org/abs/2112.06878}{4-OGC: Catalog of gravitational waves from compact-binary mergers.} pre-print arXiv: 2112.06878 (2021)
	\item[\cite{Schafer:2022dxv}] Marlin B. Schäfer, Ond\v{r}ej Zelenka, Alexander H. Nitz, He Wang, Shichao Wu, Zong-Kuan Guo, Zhoujian Cao, Zhixiang Ren, Paraskevi Nousi, Nikolaos Stergioulas, Panagiotis Iosif, Alexandra E. Koloniari, Anastasios Tefas, Nikolaos Passalis, Francesco Salemi, Gabriele Vedovato, Sergey Klimenko, Tanmaya Mishra, Bernd Brügmann, Elena Cuoco, E.A. Huerta, Chris Messenger and Frank Ohme, \href{http://arxiv.org/abs/2209.11146}{MLGWSC-1: The first Machine Learning Gravitational-Wave Search Mock Data Challenge.} pre-print arXiv: 2209.11146 (2022)
\end{itemize}